---
title: "[Aim 1] Evaluating the impact of endocrine-disrupting chemicals on breast density- An update: GEE Models "
author: "Lara Yoon"
date: "7/13/2021"
output: 
  html_document: 
    toc: true
    toc_float: true 
---

R markdown file for EDC-BD GEE modeling. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/laray/Box/Projects/2020-Dissertation/Analysis')

#In your code chunk, use echo=TRUE to display the code, message=FALSE to suppress messages, warning=FALSE to suppress warning, and include = FALSE to prevent code and results from appearing in the finished file.
```

## Setup 

```{r packages, include = FALSE}
pacman::p_load(devtools,here,BiocManager, dagitty,
               tidyverse, reshape2, janitor, knitr, kableExtra,
               tableone, readxl, corrplot, writexl,
               rmcorr, geepack, ggplot2, sjstats, psych, 
               ggpubr, ggsci, lme4, corrr, irr, ICC.Sample.Size, ICC,
               Hmisc, mice, psych, DescTools, RColorBrewer, 
               sjPlot, sjmisc, psycho, lsmeans, lmerTest, 
               emmeans, jtools, Publish)
pacman::p_loaded()


```

```{r data, message=FALSE, results = FALSE, warning=FALSE}
# Import data from file "Aim1_EDC-BD_Setup.RMD" 

load(file = "Data/FinalData/dat2.Rdata")

```

```{r dag}

# Testable implications 
testImplications <- function( covariance.matrix, sample.size ){
	library(ggm)
	tst <- function(i){ pcor.test( pcor(i,covariance.matrix), length(i)-2, sample.size )$pvalue }
tos <- function(i){ paste(i,collapse=" ") }
implications <- list(c("BD","Breast feeding","Diet","Fat%","Age","EDC","M_Edu"),
		c("BD","Birth mode","Diet","Fat%","Age","EDC","M_Edu"),
		c("M_Edu","Fat%","Breast feeding","Birth mode"),
		c("M_Edu","Diet"),
		c("M_Edu","Age"),
		c("M_Edu","Age at menarche","EDC"),
		c("Breast feeding","Birth mode","M_Edu"),
		c("Breast feeding","Diet"),
		c("Breast feeding","Age"),
		c("Breast feeding","Age at menarche","EDC"),
		c("Fat%","Age","Diet"),
		c("Fat%","Age at menarche","EDC"),
		c("Birth mode","Diet"),
		c("Birth mode","Age"),
		c("Birth mode","Age at menarche","EDC"),
		c("Diet","Age at menarche","EDC"),
		c("Age","Age at menarche","EDC"))
	data.frame( implication=unlist(lapply(implications,tos)),
		pvalue=unlist( lapply( implications, tst ) ) )

}


# Model code
g <- dagitty('dag {
bb="0,0,1,1"
"Age at menarche" [pos="0.456,0.657"]
"Birth mode" [pos="0.128,0.497"]
"Breast feeding" [pos="0.208,0.416"]
"Fat%" [adjusted,pos="0.424,0.476"]
Age [adjusted,pos="0.401,0.289"]
BD [outcome,pos="0.700,0.593"]
Diet [adjusted,pos="0.299,0.367"]
EDC [exposure,pos="0.279,0.594"]
M_Edu [adjusted,pos="0.220,0.233"]
"Age at menarche" -> BD
"Birth mode" -> "Fat%"
"Birth mode" -> EDC
"Breast feeding" -> "Fat%"
"Breast feeding" -> EDC
"Fat%" -> BD
"Fat%" -> EDC
Age -> BD
Age -> Diet
Age -> EDC
Diet -> "Fat%"
Diet -> BD
Diet -> EDC
EDC -> "Age at menarche"
EDC -> BD
M_Edu -> "Birth mode"
M_Edu -> "Breast feeding"
M_Edu -> BD
M_Edu -> EDC
}')
plot(g)

```

## GEE modeling 

The primary issue for examining exposure data with this time-varying structure is how to account for the longitudinal features of the exposure trajectory in a final disease risk model with the outcome, conditional on the complete set of exposure measures. 
This type of problem is somewhat unique, as in most regression settings the independent variable and the dependent variable are either both cross-sectional, both longitudinal, or the outcome is longitudinal with a single baseline measure of exposure.
Commonly used approaches such as multiple cross-sectional models or simultaneous multivariate logistic regression modeling suffer from problems of multiple testing or collinearity. 
Several statistical approaches exist to help identify windows of vulnerability with repeated exposure measurements (Chen et al. 2015; Sánchez et al. 2011). Sanchez et al. and Chen et al. have examined the use of GEE methods to model repeated measures of exposure with a single outcome. 

We elected to use the "Multiple Informants" approached described by Sanchez, which has the following characteristics: 

- Can test difference of estimated exposure effects across time

- At least one sample per participant (does not require one per window)

- Assumes homogeneous exposure effect within window

- Missing data assumption: MCAR 

- No restrictions on assumed time spacing between one window and another

- Not subject to collinearity problems


My code is based on SAS code adapted from Sanchez et al and was replicated in R.  Link: https://ehp.niehs.nih.gov/action/downloadSupplement?doi=10.1289%2Fehp.1002453&file=ehp.1002453.s001.acco.pdf

Other resources include Chen et al. (Statistical methods for modeling repeated measures of maternal environmental exposure biomarkers during pregnancy in association with preterm birth).
Link: https://ehjournal.biomedcentral.com/articles/10.1186/1476-069X-14-9#Sec2


### Testing code  

This section uses BP3 to test the various GEE models. We have a crude model with just interaction with Tanner stage (moments) and a full model that is adjusted for potential confounders. 


We are considering the following basic model, where: 

- Y: Breast outcome (%FGV), log-transformed, continuous

- X: EDC exposure (BP3), log-transformed, continuous

- M: Moments (B1, B4, 1YPM), categorical 3-level variable 

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_x X +  \hat{\beta}_{m1}(M=B4) + \hat{\beta}_{m2}(M=1YPM) + \hat{\beta}_{xm1}X(M=B4) + \hat{\beta}_{xm2}X(M=1YPM) $$
We can get the slopes for each category with: 

$$ slope_{X|M=2}=\beta_x+ \beta_{xm2} $$
$$ slope_{X|M=3}=\beta_x+ \beta_{xm3} $$
$$ slope_{X|M=1}=\beta_x $$

The confidence intervals for the new slopes (those that were added) are: 

$$ \sqrt{SE_{1}^2 + SE_{2}^2 +2Cov(b1,b2)}$$


Other resources: 

- ME3

- https://stats.idre.ucla.edu/r/seminars/interactions-r/#s4



```{r gee coef model 1}

set.seed(801)

## Testing model output for BP3 

# MODEL 1: Interaction model with main effects, no confounder adjustment
testfit1 <- geeglm(formula=log_pFGV ~ moments + log_bp3 + log_bp3*moments,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)

# Get beta coefficient & standard error output
coefs <- coef(summary(testfit1))

# Get var-covar matrix
covs <- as.data.frame(vcov(testfit1))

# Calculate new coefficients and standard errors accounting for interaction terms by hand
betaB1 <- coefs[4,1]
seB1 <- coefs[4,2]
betaB4 <- coefs[4,1] + coefs[5,1]
seB4 <- sqrt(coefs[4,2]^2 + coefs[5,2]^2 + 2*(covs[5,4]))
betaPM1 <- coefs[4,1] + coefs[6,1]
sePM1 <- sqrt(coefs[4,2]^2 + coefs[6,2]^2 + 2*(covs[6,4]))

# Calculate 95% confidence intervals 
cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)

# Creating a data frame 
bp3df_model1 <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
edc <- "bp3"
bp3df_model1[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)  
bp3df_model1kab <- kable(bp3df_model1, 
                   caption="Effect of EDC on %FGV by Moment (Model 1)- hand calculated ", 
                   digits=4)
bp3df_model1kab


# Testing hand-calculated SE & CI with emtrends
bp3_model1EM <- emtrends(testfit1, ~ moments, var="log_bp3")
bp3_model1EMkab <- kable(bp3_model1EM, 
                   caption="Effect of EDC on %FGV by Moment (Model 1)- function emtrends ", 
                   digits=4)
bp3_model1EMkab
```

#### Interpretation (model 1)
The beta coefficients and standard errors from Model 1 (crude model) are: 

|Parameter          | Estimate | Std.err | 
|------------------:|:--------:|:--------|
|intercept          |  1.7051  | 0.0219  |   
|momentsb4          | -0.0414  |0.0219   |
|momentspm1         | -0.0401  |0.0179   |
|log_bp3            | -0.0207  |0.0166   |
|momentsb4:log_bp3  |  0.0371  |0.0188   |  
|momentspm1:log_bp3 |  0.0304  |0.0204   |


Model: log_FGV = b0 + b1(momentsb4) + b2(momentspm1) + b3(log_bp3) + b4(momentsb4:log_bp3) + b5(momentspm1:log_bp3)

To interpret these values:

- The change in %FGV associated with a 1-unit increase in BP3 at Tanner B1: b4  =  -0.021 

- The change in %FGV associated with a 1-unit increase in BP3 at Tanner B4: b3 + b4  = 0.016

- The change in %FGV associated with a 1-unit increase in BP3 at Tanner PM1: b3 + b5 = 0.010



- b3 : At timepoint B1, a one log increase in EDC is associated with a decrease of -0.0207 log %FGV

- b4 + b5 :  The difference in log %FGV for a one log increase in EDC at timepoint B4 

- b5 + b6 : The difference in log %FGV for a one log increase in EDC at timepoint pm1


- In practical terms, for a 1 unit change in the predictor the difference in the log of Y (the outcome) is -0.02. Recalling our rules of logs (in this case we are using log 10), we can do 10^(-0.02)=0.955. Therefore, for a 1 unit increase in log bp3 (which is a 10-fold increase in bp3), we see a (1-0.955)%=.045 x100 = 4.5% decrease in percent breast density. 


#### Re-run, using model 2 (confounder adjusted)

```{r gee coef model 2}

set.seed(801)

# MODEL 2: Interaction model with main effects & adjusting for age, fat, medu, diet
testfit2 <- geeglm(formula=log_pFGV ~ moments + log_bp3 + log_bp3*moments +  medu + age*moments + fatp*moments + avgcal*moments + log_cre*moments,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)

# Get beta coefficient & standard error output
coefs <- coef(summary(testfit2))

# Get var-covar matrix
covs <- vcov(testfit2)

# Calculate new coefficients and standard errors accounting for interaction 
betaB1 <- coefs[4,1]
seB1 <- coefs[4,2]
betaB4 <- coefs[4,1] + coefs[10,1]
seB4 <- sqrt(coefs[4,2]^2 + coefs[10,2]^2 + 2*(covs[4,10]))
betaPM1 <- coefs[4,1] + coefs[11,1]
sePM1 <- sqrt(coefs[4,2]^2 + coefs[11,2]^2 + 2*(covs[4,11]))

# Calculate 95% confidence intervals
cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)

# Creating a data frame 
bp3df <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
edc <- "bp3"
bp3df[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)  
bp3df_model2kab <- kable(bp3df, 
                   caption="Effect of EDC on %FGV by Moment (Model 2)- hand calculated ", 
                   digits=4)
bp3df_model2kab



# Testing hand-calculated SE & CI with emtrends
bp3_model2EM <- emtrends(testfit2, ~ moments, var="log_bp3")
bp3_model2EMkab <- kable(bp3_model2EM, 
                   caption="Effect of EDC on %FGV by Moment (Model 2)- function emtrends ", 
                   digits=4)
bp3_model2EMkab

```

### GEE Models: %FGV
For all models we are using a log-transformed EDC value and a log-transformed breast outcome. 

```{r gee pFGV}
set.seed(801)

# List of EDCs (one model per EDC)
list.edcs <- c("log_bp3", "log_bpa", "log_bps", "log_tcs", "log_mepb", "log_etpb", "log_prpb", "log_mbp", "log_mbzp", "log_mcpp", "log_mecpp","log_mehhp",  "log_mehp", "log_meohp", "log_mep", "log_mibp", "log_dehp", "log_hiphth", "log_lophth", "log_parbf", "log_phenf")


# Function for model1 : basic model with interaction 
geefun <- function(var) {
  form <- as.formula(paste0("log_pFGV ~ 0 + moments + ",var," + moments*",var))
  fit_var <- geeglm(formula=form,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
   coefs <- coef(summary(fit_var)) 
  covs <- vcov(fit_var)
  betaB1 <- coefs[4,1]
  seB1 <- coefs[4,2]
  betaB4 <- coefs[4,1] + coefs[5,1]
  seB4 <- sqrt(coefs[4,2]^2 + coefs[5,2]^2 + 2*(covs[5,4]))
  betaPM1 <- coefs[4,1] + coefs[6,1]
  sePM1 <- sqrt(coefs[4,2]^2 + coefs[6,2]^2 + 2*(covs[6,4]))
  cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
  ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
  cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
  ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
  cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
  ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)
  vardf <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
  edc <- as.character(var)
  vardf[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)
}


# Output for model1: basic model with interaction
geeres <- list.edcs %>%  
  lapply(geefun)
geeresdf <- as.data.frame(do.call(rbind, geeres))
names(geeresdf) <- c("edc", "betaB1", "cilowerB1", "ciupperB1", "betaB4", "cilowerB4", "ciupperB4", "betaPM1", "cilowerPM1", "ciupperPM1")

geeresdf1kab <- kable(geeresdf, 
                   caption="Effect of EDC on %FGV by Moment", 
                   digits=3)
geeresdf1kab


# Function for model2 :  model with interaction & potential confounders
geefun2 <- function(var) {
  form2 <- as.formula(paste0("log_pFGV ~ 0 + moments + ",var," + moments*",var,"+ medu + age*moments + fatp*moments + avgcal*moments + log_cre*moments"))
  fit_var2 <- geeglm(formula=form2,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
  coefs <- coef(summary(fit_var2)) 
  covs <- vcov(fit_var2)
  betaB1 <- coefs[4,1]
  seB1 <- coefs[4,2]
  betaB4 <- coefs[4,1] + coefs[10,1]
  seB4 <- sqrt(coefs[4,2]^2 + coefs[10,2]^2 + 2*(covs[4,10]))
  betaPM1 <- coefs[4,1] + coefs[11,1]
  sePM1 <- sqrt(coefs[4,2]^2 + coefs[11,2]^2 + 2*(covs[4,11]))
  cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
  ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
  cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
  ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
  cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
  ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)
  vardf <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
  edc <- as.character(var)
  vardf[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)
}


# Output for model2:  model with interaction & potential confounders
geeres2 <- list.edcs %>%  
  lapply(geefun2)
geeresdf2 <- as.data.frame(do.call(rbind, geeres2))
names(geeresdf2) <- c("edc", "betaB1", "cilowerB1", "ciupperB1", "betaB4", "cilowerB4", "ciupperB4", "betaPM1", "cilowerPM1", "ciupperPM1")

geeresdf2kab <- kable(geeresdf2, 
                   caption="Effect of EDC on %FGV by Moment (adjusted for age, fat%, medu, avgcal)", 
                   digits=3)
geeresdf2kab


# Save table for later formatting 
geeresdf <- apply(geeresdf,2,as.character)
geeresdf2 <- apply(geeresdf2,2,as.character)
write.csv(geeresdf, "Output/Aim1/07132021/gee_pfgv_basic.csv")
write.csv(geeresdf2, "Output/Aim1/07132021/gee_pfgv_adj.csv")

```

### GEE Models: absolute FGV
For all models we are using a log-transformed EDC value and a log-transformed breast outcome. 

```{r gee aFGV}
set.seed(801)

# List of EDCs (one model per EDC)
list.edcs <- c("log_bp3", "log_bpa", "log_bps", "log_tcs", "log_mepb", "log_etpb", "log_prpb", "log_mbp", "log_mbzp", "log_mcpp", "log_mecpp","log_mehhp",  "log_mehp", "log_meohp", "log_mep", "log_mibp", "log_dehp", "log_hiphth", "log_lophth", "log_parbf", "log_phenf")


# Function for model1 : basic model with interaction 
geefun <- function(var) {
  form <- as.formula(paste0("log_aFGV ~ 0 + moments + ",var," + moments*",var))
  fit_var <- geeglm(formula=form,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
   coefs <- coef(summary(fit_var)) 
  covs <- vcov(fit_var)
  betaB1 <- coefs[4,1]
  seB1 <- coefs[4,2]
  betaB4 <- coefs[4,1] + coefs[5,1]
  seB4 <- sqrt(coefs[4,2]^2 + coefs[5,2]^2 + 2*(covs[5,4]))
  betaPM1 <- coefs[4,1] + coefs[6,1]
  sePM1 <- sqrt(coefs[4,2]^2 + coefs[6,2]^2 + 2*(covs[6,4]))
  cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
  ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
  cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
  ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
  cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
  ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)
  vardf <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
  edc <- as.character(var)
  vardf[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)
}

# Output for model1: basic model with interaction
geeres <- list.edcs %>%  
  lapply(geefun)
geeresdf <- as.data.frame(do.call(rbind, geeres))
names(geeresdf) <- c("edc", "betaB1", "cilowerB1", "ciupperB1", "betaB4", "cilowerB4", "ciupperB4", "betaPM1", "cilowerPM1", "ciupperPM1")

geeresdf1kab <- kable(geeresdf, 
                   caption="Effect of EDC on aFGV by Moment", 
                   digits=3)
geeresdf1kab


# Function for model2 :  model with interaction & potential confounders
geefun2 <- function(var) {
  form2 <- as.formula(paste0("log_aFGV ~ 0 + moments + ",var," + moments*",var,"+ medu + age*moments + fatp*moments + avgcal*moments + log_cre*moments"))
  fit_var2 <- geeglm(formula=form2,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
  coefs <- coef(summary(fit_var2)) 
  covs <- vcov(fit_var2)
  betaB1 <- coefs[4,1]
  seB1 <- coefs[4,2]
  betaB4 <- coefs[4,1] + coefs[10,1]
  seB4 <- sqrt(coefs[4,2]^2 + coefs[10,2]^2 + 2*(covs[4,10]))
  betaPM1 <- coefs[4,1] + coefs[11,1]
  sePM1 <- sqrt(coefs[4,2]^2 + coefs[11,2]^2 + 2*(covs[4,11]))
  cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
  ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
  cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
  ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
  cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
  ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)
  vardf <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
  edc <- as.character(var)
  vardf[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)
}


# Output for model2:  model with interaction & potential confounders
geeres2 <- list.edcs %>%  
  lapply(geefun2)
geeresdf2 <- as.data.frame(do.call(rbind, geeres2))
names(geeresdf2) <- c("edc", "betaB1", "cilowerB1", "ciupperB1", "betaB4", "cilowerB4", "ciupperB4", "betaPM1", "cilowerPM1", "ciupperPM1")

geeresdf2kab <- kable(geeresdf2, 
                   caption="Effect of EDC on aFGV by Moment (adjusted for age, fat%, medu, avgcal)", 
                   digits=3)
geeresdf2kab


# Save table for later formatting 
geeresdf <- apply(geeresdf,2,as.character)
geeresdf2 <- apply(geeresdf2,2,as.character)
write.csv(geeresdf, "Output/Aim1/07132021/gee_aFGV_basic.csv")
write.csv(geeresdf2, "Output/Aim1/07132021/gee_aFGV_adj.csv")
```

### GEE Models: Total Breast Volume
For all models we are using a log-transformed EDC value and a log-transformed breast outcome. 

```{r gee tBV}
set.seed(801)

# List of EDCs (one model per EDC)
list.edcs <- c("log_bp3", "log_bpa", "log_bps", "log_tcs", "log_mepb", "log_etpb", "log_prpb", "log_mbp", "log_mbzp", "log_mcpp", "log_mecpp","log_mehhp",  "log_mehp", "log_meohp", "log_mep", "log_mibp", "log_dehp", "log_hiphth", "log_lophth", "log_parbf", "log_phenf")


# Function for model1 : basic model with interaction 
geefun <- function(var) {
  form <- as.formula(paste0("log_tBV ~ 0 + moments + ",var," + moments*",var))
  fit_var <- geeglm(formula=form,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
   coefs <- coef(summary(fit_var)) 
  covs <- vcov(fit_var)
  betaB1 <- coefs[4,1]
  seB1 <- coefs[4,2]
  betaB4 <- coefs[4,1] + coefs[5,1]
  seB4 <- sqrt(coefs[4,2]^2 + coefs[5,2]^2 + 2*(covs[5,4]))
  betaPM1 <- coefs[4,1] + coefs[6,1]
  sePM1 <- sqrt(coefs[4,2]^2 + coefs[6,2]^2 + 2*(covs[6,4]))
  cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
  ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
  cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
  ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
  cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
  ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)
  vardf <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
  edc <- as.character(var)
  vardf[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)
}

# Output for model1: basic model with interaction
geeres <- list.edcs %>%  
  lapply(geefun)
geeresdf <- as.data.frame(do.call(rbind, geeres))
names(geeresdf) <- c("edc", "betaB1", "cilowerB1", "ciupperB1", "betaB4", "cilowerB4", "ciupperB4", "betaPM1", "cilowerPM1", "ciupperPM1")

geeresdf1kab <- kable(geeresdf, 
                   caption="Effect of EDC on tBV by Moment", 
                   digits=3)
geeresdf1kab


# Function for model2 :  model with interaction & potential confounders
geefun2 <- function(var) {
  form2 <- as.formula(paste0("log_tBV ~ 0 + moments + ",var," + moments*",var,"+ medu + age*moments + fatp*moments + avgcal*moments + log_cre*moments"))
  fit_var2 <- geeglm(formula=form2,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
  coefs <- coef(summary(fit_var2)) 
  covs <- vcov(fit_var2)
  betaB1 <- coefs[4,1]
  seB1 <- coefs[4,2]
  betaB4 <- coefs[4,1] + coefs[10,1]
  seB4 <- sqrt(coefs[4,2]^2 + coefs[10,2]^2 + 2*(covs[4,10]))
  betaPM1 <- coefs[4,1] + coefs[11,1]
  sePM1 <- sqrt(coefs[4,2]^2 + coefs[11,2]^2 + 2*(covs[4,11]))
  cilowerB1 <- betaB1 - (qnorm(0.975)*seB1)
  ciupperB1 <- betaB1 + (qnorm(0.975)*seB1)
  cilowerB4 <- betaB4 - (qnorm(0.975)*seB4)
  ciupperB4 <- betaB4 + (qnorm(0.975)*seB4)
  cilowerPM1 <- betaPM1 - (qnorm(0.975)*sePM1)
  ciupperPM1 <- betaPM1 + (qnorm(0.975)*sePM1)
  vardf <- data.frame(EDC = character(), betaB1 = numeric(), cilowerB1 = numeric(), ciupperB1 = numeric(),
                    betaB4 = numeric(), cilowerB4 = numeric(), ciupperB4 = numeric(),
                    betaPM1 = numeric(), cilowerPM1 = numeric(), ciupperPM1 = numeric(),
                    stringsAsFactors = FALSE)
  edc <- as.character(var)
  vardf[1, ] <- list(edc, betaB1, cilowerB1, ciupperB1, betaB4, cilowerB4, ciupperB4, betaPM1, cilowerPM1, ciupperPM1)
}


# Output for model2:  model with interaction & potential confounders
geeres2 <- list.edcs %>%  
  lapply(geefun2)
geeresdf2 <- as.data.frame(do.call(rbind, geeres2))
names(geeresdf2) <- c("edc", "betaB1", "cilowerB1", "ciupperB1", "betaB4", "cilowerB4", "ciupperB4", "betaPM1", "cilowerPM1", "ciupperPM1")

geeresdf2kab <- kable(geeresdf2, 
                   caption="Effect of EDC on tBV by Moment (adjusted for age, fat%, medu, avgcal)", 
                   digits=3)
geeresdf2kab


# Save table for later formatting 
geeresdf <- apply(geeresdf,2,as.character)
geeresdf2 <- apply(geeresdf2,2,as.character)
write.csv(geeresdf, "Output/Aim1/07132021/gee_tBV_basic.csv")
write.csv(geeresdf2, "Output/Aim1/07132021/gee_tBV_adj.csv")
```

## Interaction 
Using ANOVA to get p-values for interaction between EDC & moments. 


```{r gee typeiii}

set.seed(801)
summary(testfit1)
summary(testfit2)

# Test the difference in exposure association across time windows with type III test

# Model 1 (with BP3)
aov1 <- anova(testfit1, test="F")
aov1_tidy <- as.data.frame(tidy(anova(testfit1, test="F")))
aov1_tidy

# Model 2 (with BP3)
aov2 <- anova(testfit2, test="F")
aov2_tidy <- as.data.frame(tidy(anova(testfit2, test="F")))
aov2
```

#### Interpretation
We also see that the interaction between moments & log_bp3 is NOT significant in the crude (p=0.13) and adjusted (p=0.092), suggesting no time-varying associations. 


```{r gee interaction, warning=FALSE}
set.seed(801)
# Function test the difference in exposure association across time windows with type III test
## pFGV 
geeint1 <- function(var) {
  form1 <- as.formula(paste0("log_pFGV ~ 0 + moments + ",var," + moments*",var,"+ medu + age*moments + fatp*moments + avgcal*moments + log_cre*moments"))
  aov_var1 <- geeglm(formula=form1,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
  aovtidy_var1 <- as.data.frame(tidy(anova(aov_var1, test="F")))
  aov_sub <- aovtidy_var1[8, c(1,4)]
} 
geeintres1 <- list.edcs %>%  
  lapply(geeint1)
geeintresbind1 <- bind_rows(geeintres1)
geeintres1kab <- kable(geeintresbind1, 
                   caption="Significance of interaction between EDC & Tanner Stage on pFGV ", 
                   digits=3)
geeintres1kab

## aFGV 
geeint2 <- function(var) {
  form2 <- as.formula(paste0("log_aFGV ~ 0 + moments + ",var," + moments*",var,"+ medu + age*moments + fatp*moments + avgcal*moments + log_cre*moments"))
  aov_var2 <- geeglm(formula=form2,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
  aovtidy_var2 <- as.data.frame(tidy(anova(aov_var2, test="F")))
  aov_sub <- aovtidy_var2[8, c(1,4)]
} 
geeintres2 <- list.edcs %>%  
  lapply(geeint2)
geeintresbind2 <- bind_rows(geeintres2)
geeintres2kab <- kable(geeintresbind2, 
                   caption="Significance of interaction between EDC & Tanner Stage on aFGV ", 
                   digits=3)
geeintres2kab

## tBV 
geeint3 <- function(var) {
  form3 <- as.formula(paste0("log_tBV ~ 0 + moments + ",var," + moments*",var,"+ medu + age*moments + fatp*moments + avgcal*moments + log_cre*moments"))
  aov_var3 <- geeglm(formula=form3,
               id=chear_pid, 
               data=dat2, 
               family=gaussian(link = "identity"), 
               corstr="independence", 
               waves=moments)
  aovtidy_var3 <- as.data.frame(tidy(anova(aov_var3, test="F")))
  aov_sub <- aovtidy_var3[8, c(1,4)]
} 
geeintres3 <- list.edcs %>%  
  lapply(geeint3)
geeintresbind3 <- bind_rows(geeintres3)
geeintres3kab <- kable(geeintresbind3, 
                   caption="Significance of interaction between EDC & Tanner Stage on tBV ", 
                   digits=3)
geeintres3kab


# Save table for later formatting 
write.csv(geeintresbind1, "Output/Aim1/07132021/gee_pfgv_adj_int.csv")
write.csv(geeintresbind2, "Output/Aim1/07132021/gee_afgv_adj_int.csv")
write.csv(geeintresbind3, "Output/Aim1/07132021/gee_tbv_adj_int.csv")

```